---
title: "QC: UHN Tsao Lung PDX XevaSet"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    df_print: paged
date: "`r format(Sys.Date())`"
---

<style>
.scroll-box {
  max-height: 350px;
  overflow-y: auto;
  background-color: inherit;
}
</style>

## Download required packages

```{r chunk-1}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("Xeva", quietly = TRUE))
    BiocManager::install("Xeva", version = "3.20")
```

## Load library

```{r chunk-2}
library(Xeva)
```

## Read Dataset

```{r chunk-3}
x.set <- readRDS("../expected_results/xevaset/UHN_Tsao_Lung_DrugResponse_2022_v1.rds")
```

```{r chunk-4}
getwd()
```

## 1. Dataset Structure and General Information

### 1.1 Dataset annotation

```{r chunk-5}
x.set
```

### 1.2 Dataset structure

```{r chunk-6}
slotNames(x.set)
```

### 1.3 Model data structure

```{r chunk-7}
head(x.set@model)
```

### 1.4 Drug data structure

```{r chunk-8}
head(x.set@drug)
```

### 1.5 Drug screening data structure

```{r chunk-9, class.output="scroll-box"}
head(x.set@sensitivity)
```

### 1.6 Batch information

```{r chunk-10, class.output="scroll-box"}
head(x.set@expDesign)
```

### 1.7 Molecular Profiles

```{r chunk-11}
head(x.set@molecularProfiles)
```

### 1.8 Mapping between drug experiment (model.id) to molecular profile (biobase.id)

```{r chunk-12}
head(x.set@modToBiobaseMap)
```

## 2. Model Data

```{r chunk-13}
# access model data
x.mod <- modelInfo(x.set)
dim(x.mod)
```

```{r chunk-14}
x.mod[1:4, ]
```

## 3. Experiment Data

```{r chunk-15}
# access experiment for a model
model.data <- getExperiment(x.set, model.id='f1.PHLC402.TPX.crizotinib.P601')
head(model.data)
```

```{r chunk-16}
model.data <- getExperiment(x.set, model.id='f1.PHLC402.TPX.control.P603')
head(model.data)
```

## 3. Batch Data

```{r chunk-17}
batch.name <- batchInfo(x.set)
batch.name[1:4]
```

```{r chunk-18}
batchInfo(x.set, batch="f1.PHLC402.TPX.crizotinib")
```

## 4. PDX Growth Cure

```{r chunk-19}
# set color of lines, error visualization with errorbar
plotPDX(x.set, batch = "f4.PHLC191.TPX.palbociclib", 
        control.col = "#a6611a", 
        treatment.col = "#018571",
        major.line.size = 1,
        max.time = 100,
        SE.plot='errorbar')
```

```{r chunk-20}
# normalize volume, change color of lines, error visualization with ribbon
plotPDX(x.set, batch = "f4.PHLC191.TPX.palbociclib", 
        vol.normal = TRUE, 
        control.col = "#a6611a", 
        treatment.col = "#018571",
        major.line.size = 1,
        max.time = 100,
        SE.plot='ribbon')
```

