---
title: "QC Report: UHN Tsao Lung PDX XevaSet"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: cosmo
    mathjax: null
---

This document is a curation QC report for the UHN Tsao Lab lung cancer PDX
`XevaSet`, built by the [Xevacurate Snakemake
pipeline](https://github.com/BHKLAB-DataProcessing/Xevacurate_Snakemake_UHN_Lung).

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(Biobase)
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(jsonlite)
  library(knitr)
  library(patchwork)
  library(scales)
  library(tidyr)
  library(Xeva)
})

theme_set(theme_minimal(base_size = 12))

repo_path <- function(...) file.path("..", ...)
```

## Load The Dataset

The curated object is stored as an `RDS` file in `expected_results/xevaset/`.

```{r load-data}
xeva_path <- repo_path(
  "expected_results", "xevaset",
  "UHN_Tsao_Lung_DrugResponse_2022_v1.rds"
)

x <- readRDS(xeva_path)
x
```

The `XevaSet` bundles model metadata, drug metadata, tumor growth
measurements, response summaries, and molecular profiles into one object.

```{r dataset-overview}
model_df <- modelInfo(x) |>
  as_tibble()
drug_df <- x@drug |>
  as_tibble()
batch_ids <- batchInfo(x)
model_metrics <- x@sensitivity$model |>
  as.data.frame(stringsAsFactors = FALSE)
batch_metrics <- x@sensitivity$batch |>
  as.data.frame(stringsAsFactors = FALSE)
experiment_csv <- fread(repo_path("results", "build", "experiment.csv"))
example_experiment <- getExperiment(x, model.id = model_df$model.id[[1]])

overview_tbl <- tibble(
  quantity = c(
    "Model records",
    "Patients",
    "Treatment-control batches",
    "Treatments including control",
    "Tumor volume measurements",
    "Molecular profile types"
  ),
  value = c(
    nrow(model_df),
    n_distinct(model_df$patient.id),
    length(batch_ids),
    n_distinct(drug_df$drug.id),
    nrow(experiment_csv),
    length(x@molecularProfiles)
  )
)

kable(overview_tbl)
```

## What Is Included?

```{r object-contents}
contents_tbl <- tibble(
  component = c(
    "Model metadata",
    "Treatment metadata",
    "Model-level response summaries",
    "Batch-level response summaries",
    "Molecular profiles",
    "Raw drug-screen files scanned",
    "Raw records excluded (unmatched drug label)"
  ),
  rows_or_items = c(
    nrow(model_df),
    nrow(drug_df),
    nrow(model_metrics),
    nrow(batch_metrics),
    length(x@molecularProfiles),
    NA_integer_,
    NA_integer_
  ),
  fields_or_summary = c(
    paste(colnames(model_df), collapse = ", "),
    paste(colnames(drug_df), collapse = ", "),
    paste(colnames(model_metrics), collapse = ", "),
    paste(colnames(batch_metrics), collapse = ", "),
    paste(names(x@molecularProfiles), collapse = ", "),
    "see Raw Data Ingestion QC below",
    "see Treatment Metadata below"
  )
)

kable(contents_tbl)
```

```{r growth-curve-table-example}
kable(head(example_experiment, 8))
```

## Raw Data Ingestion QC

Every raw drug-screen Excel file is scanned and parsed before curation. Files
that fail to parse (e.g. an unexpected sheet layout) are logged rather than
silently dropped.

```{r qc-report-summary}
qc_report <- fromJSON(repo_path("results", "process", "qc_report.json"))

qc_summary <- qc_report |>
  count(issue, name = "files", sort = TRUE)

kable(qc_summary)
```

```{r qc-report-issues}
problem_files <- qc_report |>
  filter(issue != "file processed successfully") |>
  mutate(file = basename(file))

if (nrow(problem_files) > 0) {
  kable(problem_files)
} else {
  kable(tibble(status = "All scanned files parsed successfully."))
}
```

## Model Metadata

Each row in the model metadata represents one drug-response experiment arm
(one mouse/passage under one treatment). `parental.model.id` links related
arms back to the same engrafted tumor line.

```{r model-metadata-head}
kable(head(model_df, 8))
```

Key metadata columns include:

* `model.id`: unique drug-experiment model identifier (`f<batch>.<modelID>.<engraftment>.<drug>.<mouse>`).
* `patient.id`: patient-level identifier shared across related arms.
* `drug`: normalized treatment name (`control` for untreated arms).
* `oncotree.code` / `primary.tumor.site` / `cancer.type.detailed`: clinical annotation.
* `model.type` / `parental.model.id`: parental vs. derivative model lineage.

```{r clinical-completeness}
clinical_columns <- c(
  "sex", "oncotree.code", "primary.tumor.site",
  "cancer.type.detailed", "TNM.stage"
)

clinical_completeness <- tibble(
  field = clinical_columns,
  models_annotated = sapply(
    clinical_columns,
    function(col) sum(!is.na(model_df[[col]]) & model_df[[col]] != "")
  ),
  models_missing = sapply(
    clinical_columns,
    function(col) sum(is.na(model_df[[col]]) | model_df[[col]] == "")
  )
) |>
  mutate(pct_annotated = round(100 * models_annotated / nrow(model_df), 1))

kable(clinical_completeness)
```

## Treatment Metadata

Treatment metadata is annotated with PubChem cross-references where a
compound identifier could be resolved.

```{r treatment-metadata}
kable(
  drug_df |>
    select(
      drug.id,
      drugname.standardized,
      target,
      dose,
      dose.unit,
      treatment.type,
      CID,
      MolecularFormula,
      InChIKey
    )
)
```

Raw drug-screen rows whose treatment label could not be matched to a
standardized drug are excluded during curation rather than guessed at.

```{r unmatched-treatment-report}
unmatched <- fread(repo_path("results", "update", "drug_filtered_out.tsv"))

kable(tibble(
  report = "Excluded rows (unmatched treatment label)",
  rows = nrow(unmatched),
  source_files = n_distinct(unmatched$file_name),
  models_affected = n_distinct(unmatched$modelID)
))
```

```{r unmatched-treatment-table}
if (nrow(unmatched) > 0) {
  unmatched |>
    count(file_name, drug, name = "rows") |>
    arrange(desc(rows)) |>
    kable()
} else {
  kable(tibble(status = "All raw treatment labels matched a standardized drug."))
}
```

## Molecular Profiles

The molecular profiles are stored as `ExpressionSet` objects in the
`XevaSet`. The matrix dimensions below are the feature-by-sample dimensions
for each profile.

```{r molecular-profile-dimensions}
profile_summary <- lapply(
  names(x@molecularProfiles),
  function(profile_name) {
    profile <- x@molecularProfiles[[profile_name]]
    feature_data <- Biobase::fData(profile)
    data.frame(
      profile = profile_name,
      features = nrow(exprs(profile)),
      samples = ncol(exprs(profile)),
      feature_id_example = rownames(exprs(profile))[[1]],
      feature_metadata = paste(colnames(feature_data), collapse = ", "),
      stringsAsFactors = FALSE
    )
  }
) |>
  bind_rows()

kable(profile_summary)
```

Feature annotation (Ensembl gene ID, Entrez ID, gene biotype) is resolved via
`biomaRt`. A small fraction of HGNC symbols fail to resolve to a current
Ensembl gene ID, most often withdrawn or ambiguous symbols.

```{r molecular-feature-annotation}
gene_feature_summary <- lapply(
  names(x@molecularProfiles),
  function(profile_name) {
    feature_data <- Biobase::fData(x@molecularProfiles[[profile_name]])
    mapped <- !is.na(feature_data$ensembl_gene_id) &
      feature_data$ensembl_gene_id != ""
    tibble(
      profile = profile_name,
      features = nrow(feature_data),
      mapped_features = sum(mapped),
      unmapped_features = sum(!mapped),
      mapped_fraction = round(mapped_features / features, 4)
    )
  }
) |>
  bind_rows()

kable(gene_feature_summary)
```

The model-to-molecular map links drug-experiment model records to omics
sample IDs. Coverage is necessarily partial: most patients were never
biopsied for sequencing, only drug-screened.

```{r molecular-map-coverage}
mod_map <- fread(repo_path("results", "build", "modToBiobaseMap.csv"))

map_by_datatype <- mod_map |>
  distinct(model.id, biobase.id, mDataType) |>
  count(mDataType, name = "model_records_mapped") |>
  arrange(mDataType)

case_coverage <- tibble(
  patients = n_distinct(model_df$patient.id),
  patients_with_rnaseq = n_distinct(
    mod_map$biobase.id[mod_map$mDataType == "RNASeq"]
  ),
  patients_with_mutation = n_distinct(
    mod_map$biobase.id[mod_map$mDataType == "mutation"]
  ),
  patients_with_cnv = n_distinct(
    mod_map$biobase.id[mod_map$mDataType == "CNV"]
  )
)

kable(map_by_datatype)
kable(case_coverage)
```

## Response Summaries

The pipeline computes model-level and batch-level response metrics using the
`Xeva` package (slope, AUC, mRECIST, TGI-derived batch metrics).

```{r model-response-availability}
model_metric_summary <- tibble(
  metric = colnames(model_metrics)[colnames(model_metrics) != "model.id"],
  available_models = colSums(!is.na(model_metrics[
    colnames(model_metrics) != "model.id"
  ])),
  pct_models = round(100 * colMeans(!is.na(model_metrics[
    colnames(model_metrics) != "model.id"
  ])), 1)
)

kable(model_metric_summary)
```

```{r batch-response-availability}
batch_metric_summary <- tibble(
  metric = colnames(batch_metrics)[colnames(batch_metrics) != "batch.name"],
  available_batches = colSums(!is.na(batch_metrics[
    colnames(batch_metrics) != "batch.name"
  ])),
  pct_batches = round(100 * colMeans(!is.na(batch_metrics[
    colnames(batch_metrics) != "batch.name"
  ])), 1)
)

kable(batch_metric_summary)
```

### Response Metric Distributions

```{r response-distributions, fig.height=6}
auc_df <- model_metrics |>
  select(AUC) |>
  filter(!is.na(AUC), AUC > 0)

slope_df <- model_metrics |>
  select(slope) |>
  filter(!is.na(slope), between(slope, -90, 90))

bar_df <- model_metrics |>
  select(best.average.response) |>
  filter(!is.na(best.average.response))

base_hist_theme <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 12),
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  )

p_auc <- ggplot(auc_df, aes(x = AUC)) +
  geom_histogram(bins = 30, fill = "#3b82b8", color = "white") +
  scale_x_continuous(labels = label_number(big.mark = ",")) +
  labs(title = "AUC", x = "AUC", y = "Models") +
  base_hist_theme

p_slope <- ggplot(slope_df, aes(x = slope)) +
  geom_histogram(bins = 30, fill = "#3b82b8", color = "white") +
  labs(title = "Slope", x = "Angle (degrees)", y = "Models") +
  base_hist_theme

p_bar <- ggplot(bar_df, aes(x = best.average.response)) +
  geom_histogram(bins = 30, fill = "#3b82b8", color = "white") +
  labs(
    title = "Best Average Response",
    x = "Percent change from baseline",
    y = "Models"
  ) +
  base_hist_theme

(p_auc + p_slope + p_bar) +
  plot_layout(ncol = 3)
```

`AUC`, `slope`, and `best.average.response` summarize different aspects of
tumor growth. Lower `best.average.response` values indicate stronger
inhibition or shrinkage relative to baseline; the long right tail on
`best.average.response` (up to `r round(max(bar_df$best.average.response))`%)
reflects a small number of rapidly-progressing control/resistant arms and is
worth spot-checking rather than assuming it's a data error.

## mRECIST Summary

The `mRECIST` matrix summarizes response by treatment and patient.

```{r mrecist-table}
mrecist_matrix <- summarizeResponse(
  x,
  response.measure = "mRECIST",
  group.by = "patient.id"
)

dim(mrecist_matrix)
```

```{r mrecist-preview}
kable(
  mrecist_matrix[seq_len(min(8, nrow(mrecist_matrix))),
    seq_len(min(8, ncol(mrecist_matrix))),
    drop = FALSE
  ]
)
```

```{r mrecist-plot, fig.width=11, fig.height=6}
plotmRECIST(
  mrecist_matrix,
  control.name = "control",
  name = "Lung",
  sort = FALSE
)
```

## Example Growth Curve

The batch-level plot pairs a treatment arm with its control arm for the same
patient.

```{r batch-example}
sample_batch <- batch_ids[grepl("palbociclib", batch_ids)][[1]]
batch_info <- batchInfo(x, batch = sample_batch)[[1]]

kable(tibble(
  field = c("Batch", "Control arms", "Treatment arms"),
  value = c(
    batch_info$batch.name,
    paste(batch_info$control, collapse = ", "),
    paste(batch_info$treatment, collapse = ", ")
  )
))
```

```{r plot-batch-growth, fig.height=5}
plotPDX(
  x,
  batch = sample_batch,
  vol.normal = TRUE,
  max.time = 100
)
```

## Waterfall Plot

Waterfall plots rank models from strongest response to weakest response for
a given treatment.

```{r waterfall-drug}
waterfall_drug <- model_df |>
  filter(drug != "control") |>
  count(drug, sort = TRUE) |>
  slice(1) |>
  pull(drug)

waterfall_drug
```

### Model-level view

`waterfall()`'s own default is `group.by = "model.id"` (not
`"patient.id"`, which is `summarizeResponse()`'s default elsewhere in this
report), so each bar below is one individual model record, not one per
patient. Patients contribute very unequal numbers of records for
`r waterfall_drug` — see the per-patient count in the next panel — so a
handful of heavily-replicated patients dominate this view.

```{r waterfall-plot-model, fig.width=10, fig.height=5}
waterfall(
  x,
  drug = waterfall_drug,
  res.measure = "best.average.response"
)
```

### Patient-level view

Collapsing to one bar per patient (`group.by = "patient.id"`) requires a
`summary.stat` other than the default `";"` — cells with more than one
record per patient would otherwise be joined into a non-numeric string and
`waterfall()` would error. Using `"mean"` gives every patient equal visual
weight regardless of how many replicate batches they contributed, at the
cost of hiding within-patient spread.

```{r waterfall-patient-counts}
model_df |>
  filter(drug == waterfall_drug) |>
  count(patient.id, name = "records") |>
  arrange(desc(records)) |>
  kable()
```

```{r waterfall-plot-patient, fig.width=10, fig.height=5}
waterfall(
  x,
  drug = waterfall_drug,
  res.measure = "best.average.response",
  group.by = "patient.id",
  summary.stat = "mean",
  title = paste(waterfall_drug, "— mean best.average.response per patient")
)
```

## Session Info

```{r session-info}
sessionInfo()
```
